%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 0_simulat.tex                                                               %
% by: David A. Clarke                                                         %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


% Normal preamble stuff.
\documentclass[12pt]{book}
\usepackage[title]{appendix}
\usepackage{mathtools,tensor,physics,slashed,textcomp,mhchem}
\usepackage{microtype} % should help with hyphenation
\usepackage{bbm} % 4-vectors

% This section is for making C++ code.
\usepackage[dvipsnames]{xcolor}
\usepackage{listings}
\lstset{
  backgroundcolor=\color{gray!10},
  basicstyle=\ttfamily,
  breakatwhitespace=false,
  breaklines=true,
  captionpos=b,
  columns=fullflexible,
  commentstyle=\color{OliveGreen},
  extendedchars=true,
  frame=single,
  keepspaces=true,
  keywordstyle=\color{blue},
  language=c++,
  numbers=none,
  numbersep=5pt,
  numberstyle=\time\color{blue},
  rulecolor=\color{Gray},
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  stepnumber=5,
  stringstyle=\color{Orchid},
  tabsize=4,
  title=\lstname
}


% davitex and required packages
\usepackage{amsmath,amssymb,graphicx,multirow,tabularx}
\usepackage{davitex}

\newcommand{\fvec}[1]{\mathbbm{#1}}
\newcommand{\units}[1]{\,[\text{#1}]}


% If you click on citations or equation references, you will be take
% to the corresponding page. 
\usepackage{hyperref}
\hypersetup{
  colorlinks=true,
  linkcolor=black,
  citecolor=magenta,
  urlcolor=blue,
}


% Put references at the end of each chapter, directly after the
% end of the chapter, call them "References" instead of
% "Bibliography", and use numbers with square brackets.
\usepackage{url,natbib,chapterbib}
\let\oldbibliography\bibliography
\renewcommand{\bibliography}[1]{{%
  \let\chapter\section%
  \oldbibliography{#1}}}
\renewcommand\bibname{References}
\setcitestyle{numbers,square}


% Make index and let it appear in table of contents.
\usepackage{makeidx}
\makeindex
\usepackage[totoc]{idxlayout}


% Prevent hyphenation.
\setlength\parindent{15pt}
\overfullrule=2cm


% Theorems and stuff.
\usepackage{amsthm}
\usepackage[many]{tcolorbox}
\tcbuselibrary{theorems}
\newtcbtheorem[number within=section]{theorem}{Theorem}{%
  breakable,
  enhanced,
  before skip=10pt,
  colback=green!5,
  colframe=green!35!black,
  fonttitle=\bfseries}{thm}
\newtcbtheorem[number within=section]{proposition}{Proposition}{%
  breakable,
  enhanced,
  before skip=10pt,
  colback=green!5,
  colframe=green!35!black,
  fonttitle=\bfseries}{prp}
\newtcbtheorem[number within=section]{lemma}{Lemma}{%
  breakable,
  enhanced,
  before skip=10pt,
  colback=green!5,
  colframe=green!35!black,
  fonttitle=\bfseries}{lem}
\newtcbtheorem[number within=section]{corollary}{Corollary}{%
  breakable,
  enhanced,
  before skip=10pt,
  colback=green!5,
  colframe=green!35!black,
  fonttitle=\bfseries}{cor}
\newtcbtheorem[number within=section]{example}{Example}{%
  breakable,
  enhanced,
  before skip=10pt,
  colback=blue!5,
  colframe=blue!35!black,
  fonttitle=\bfseries}{cor}
\theoremstyle{definition}
\newtheorem*{defn}{Definition}


% Some environments for some front matter.
\newenvironment{frontstuff}
  {\centering\chapter*{}}
  {\clearpage}
\newenvironment{listofabbrev}
  {\chapter*{List of Abbreviations}}
  {\clearpage}


% Feynman rule references
\definecolor{light}{gray}{.85}
\newcommand{\F}[1]{\colorbox{light}{{F}{#1}}}


% Note that book class by default is formatted to be printed back-to-back.
\title{\bf Simulating Reality: A First Encounter with
Lattice Field Theory for First-Year Undergraduates} 
\author{David Clarke}


\begin{document}
\frontmatter                            % roman page no., suppress chapter
\maketitle                              % print title page


\begin{frontstuff} % ----------------------------------------------- FRONT STUFF

\section*{Symbology}
\begin{tabular}{ll}
$\exists$       & There exists \\
$\forall$       & For all \\
$\in$           & Is a member of the set \\
$\leq$          & Less than or equal to; is a subgroup of\\
$\approx$       & Approximate equality \\
$\equiv$        & Is defined as \\
$\propto$       & Is proportional to \\

$a$             & Lattice spacing \\
$\C$            & Complex numbers \\
$c$             & Speed of light\\
$G$             & Newton's gravitational constant\\
$\hbar$         & Planck's constant \\
$k_B$           & Boltzmann's constant \\
$\log$          & Natural logarithm \\
$\N$            & Natural numbers \\
$N_c$           & Number of colors \\
$N_s$           & Lattice extension in a spatial direction \\
$N_\tau$        & Lattice extension in temperature/Euclidean time direction\\
$\pr{X}$        & Probability of event $X$ \\
$\Q$            & Rational numbers \\
$q$             & Gaussian or Student difference test \\
$\R$            & Real numbers \\
$\SU(N)$        & Special unitary group of degree $N$ \\
$\sigma$        & String tension\\
$\sigma_i$      & A Pauli matrix\\
$T_c$           & Deconfining phase transition temperature\\
$\mathbb{Z}$    & Integers \\


\end{tabular}
\clearpage



\section*{Abbreviations}
\begin{tabular}{ll}
BC      &       Boundary condition \\
CCW     &       Counterclockwise\\
CDF     &       Cumulative distribution function \\
CLT     &       Central limit theorem\\
CW      &       Clockwise\\
EM      &       Electromagnetic\\
FLOP    &       Floating point operation\\
LFT     &       Lattice field theory\\
LGT     &       Lattice gauge theory\\
LQCD    &       Lattice QCD\\
LHS     &       Left hand side\\
LLN     &       Law of large numbers\\
MCMC    &       Markov chain Monte Carlo\\
OOP     &       Object-oriented programming \\
PDF     &       Probability distribution function \\
QCD     &       Quantum chromodynamics \\
QM      &       Quantum mechanics \\
RHS     &       Right hand side \\
RNG     &       Random number generator \\
SM      &       Standard Model \\
s.t.    &       Such that \\
UV      &       Ultraviolet \\
WLOG    &       Without loss of generality
\end{tabular}
\clearpage



\end{frontstuff} % ----------------------------------------------------- PREFACE


\chapter{Acknowledgements}
These notes were developed in part with helpful feedback from students
and friends who had a look at what I wrote.
In particular I would like to thank Grant Curell and my students Kai Ebira
and Daeton McClure for suggestions to help improve the readability of
the notes and make the conveyed information more complete.
Grant especially devoted much time to helping make the discussion about
matrices and multivariable calculus in Chapter 1 more pedagogical.


\chapter{Preface}

Particle physics is the subdiscipline of physics that studies the smallest
particles in existence. Some of these particles are the building blocks of all known
matter, and others are physical manifestations of forces; indeed from the
modern perspective of particle physics, fundamental forces are mediated by
particles called bosons. This is the most fundamental level of nature we
experimentally understand, so in that sense, particle physics studies
the most basic building blocks of reality.

Lattice field theory (LFT) or lattice gauge theory (LGT) is a framework that
allows us to study particle physics using a computer. There are several
strategies for doing particle physics calculations on the market. In some cases
LFT serves as a cross-check of these other approaches. Sometimes, LFT can be
used to compute physical quantities that can't be computed using other methods.

Lattice calculations generate random snapshots of space-time, then measure a
physical quantity, such as a particle mass, on those snapshots. We then
perform statistical analyses on these measurements. Generating the snapshots is 
highly computationally demanding, utilizing in some cases a significant 
fraction of the resources of the most powerful supercomputers in the world. 
Therefore a significant portion of LFT work is focused on high-performance
computing.

The intention of these notes is to give an introduction to LFT that is
readable by a first-year undergraduate in physics. To get a full understanding
of LFT will not be possible for a scientist at this stage in their career, since
LFT lies at the intersection of many fields of math, physics, and computer
science, many of which one will not see until graduate school. Still, my hope is
that the reader can get a heuristic understanding of the field, at least enough
to be able to write scripts that carry out rudimentary statistical analyses on
some physical quantities while having some intuition of what it means.

I attempted to make this text relatively self-contained. However there will be
many parts that I do not explain in detail and must be taken for granted.
Moreover I assume the reader already has had a course with 
\begin{itemize}
  \item calculus (limits, derivatives, and integrals)
  \item energy and momentum conservation
  \item forces
\end{itemize}
Moreover, it is helpful if the reader already knows just a little bit about programming.
This text is developed in tandem with a course that gives first-year
undergraduates some experience in particle physics research, and in that course,
Python is the language of choice.

The layout is pretty simple: There are some preliminary chapters with
hopefully enough additional math and physics to give you a foothold in the lattice
framework. I would recommend not to skip anything in these chapters, even if it
includes things you already know, just so that you can get used to my notation
and so that you can gain a little context why these topics are of interest.
The next chapter is an introduction to the ideas and history of LFT.

One way to gain some familiarity with lattice QCD is to read through these
notes from start to finish. If you are using these notes in the context
of an undergraduate research experience, you can also use it just as
reference material. That is, you can open up a chapter or go to an index
to try to gain some information about a topic or some jargon that is
unfamiliar to you.

\tableofcontents                        % print table of contents
\mainmatter                             % arabic page no., include chapter


% These are the chapters of the book.
\include{ch_math}
\include{ch_physics}
\include{ch_LFT}


\printindex


\end{document}
